Equalizer 2.0\\Convergence of a Parallel Rendering Framework

### 2017-04-03 paperIEEE.tex
Improve object section
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@25 db7c848b-7755-4df2-b763-8796797afa52

UDP-based reliability protocol

@@

Adapting an existing application for parallel rendering requires the
synchronization of application data across the processes in the parallel
rendering setup. Existing parallel rendering frameworks address this often
poorly, at best they rely on MPI to distribute data. Real-world, interactive
visualization applications are typically written in C++ and have complex data
models and class hierarchies to represent their application state. As outlined
in \cite{EMP:09}, the parallel rendering code in an Equalizer application only
needs access to the data needed for rendering, as all application logic is
centralized in the application main thread. We have encountered two main
approaches to address this distribution: Using a shared filesystem for static
data and using data distribution for static and dynamic data.

Distributed objects in Collage provide powerful, object-oriented data
distribution for C++ objects. They facilitate the implementation of data
distribution in a cluster environment. Distributed objects are created by
subclassing from \textsf{co::Serializable} or \textsf{co::Object}. The
application programmer implements serialization and deserialization of the
distributed data. Distributed objects can be static (immutable) or
dynamic. Objects have a universally unique identifier (UUID) to address them
cluster-wide. A master-slave model is used to establish mapping and data
synchronization across processes. Typically, the application mail loop registers
a master instance and communicates the UUID to the render clients, which map
their instance to the given identifier. The following object types are
available:

@@

Besides the instance data serialization methods used to map an object, versioned
objects may implement \textsf{pack} and \textsf{unpack} to serialize or
deserialize the changes since the last version.

@@

@@

\subsection{Raster}
\subsection{Omegalib}

@@

Distribute large ply model to 1..16 nodes using TCP, IB and RSP

>>>>

RSP is an implementation of a reliable multicast protocol over unreliable UDP
transport. RSP behaves similarly to TCP, it provides full reliability and
ordering of the data, and slow receivers will eventually throttle the
sender. This behaviour is needed to guarantee delivery of data in all
situations. Pragmatic multicast (PGM~\cite{pgm}) provides full ordering, but
slow clients will disconnect from the multicast session instead of throttling
the send rate.

RSP is optimized for LAN performance. In the following we will outline the RSP
protocol and implementation and motivate the design decisions. Any defaults
given below are for Linux or OS X, the Windows UDP stack requires different
default values which can be found in the implementation.

Our RSP implementation uses a separate protocol thread for each RSP connection,
which handles all reads and writes on the multicast group. This thread
implements the protocol handling and communicates with the application threads
through thread-safe queues, as shown in \fig{fRSPProto}. These queues contain
datagrams, which are prefixed by a type-specific header. Each connection has a
configurable number of buffers, (1024 by default) of a configurable MTU (1470
bytes default) which are either free for use or in-use for reading or writing.

%%     1,     // RSP_ERROR_DOWNSCALE
%%     5,     // RSP_ERROR_UPSCALE
%%     20,    // RSP_ERROR_MAXSCALE
%%     3,     // RSP_MIN_SENDRATE_SHIFT

%%     17,    // RSP_ACK_FREQUENCY

\begin{figure}[ht]\center
  \includegraphics[width=\columnwidth]{images/rspPackets}
  \caption{\label{fRSPProto}Reliable Stream Protocol}
\end{figure}

A client opens a listening multicast RSP connection. Similar to TCP, this
listener will create one connected RSP connection for each remote member in the
multicast group, that is, on each node $n$ RSP connection instances exist, $1$
for the local listener and $n-1$ for all other nodes. A multicast group
therefore looks like a fully connected set of TCP connections between $n$ nodes
for the user.

Each member of a multicast group has a group-unique 16 bit identifier, which is
randomly selected, announced to the group, and if no collisions are notified by
any other member, confirmed for this connection. This mechanism allows
addressing of connections with a low protocol overhead. The datagrams used in
this exchange contain no payload, and they contain two bytes type (HELLO,
HELLO\_REPLAY, HELLO\_DENY, HELLO\_CONFIRM), two bytes protocol version and two
bytes connection identifier.

The \textsf{\_appBuffers} queue, which is blocking, contains received UDP
datagrams on incoming connections, or empty buffers for writing on outgoing
connections. The \textsf{\_threadBuffers} queue is lock-free and used by the
protocol thread to store empty buffers on incoming connections and pre-packaged
UDP datagrams on outgoing connections. The application thread pushes buffers
into this queue, consumed by the protocol thread. A datagram is up to 64
kilobytes (UDP limit, configurable) and contains a header (two bytes type, two
bytes size, two bytes source identifier, two bytes sequence number) and its
payload. The sequence number is used for positive (ack) and negative (nack)
acknowledgments and reordering of packets.

The \textsf{\_writeBuffers std::deque} is only used by the protocol
thread. It contains already sent, but not yet acknowledged datagrams. Its
datagrams are resent upon reception of a nack or recycled into
\textsf{\_appBuffers} when acknowledged by all receivers. Similarly, the
\textsf{\_recvBuffers std::deque} contains received out-of-order packets already
received by an incoming connection. Upon each reception it is checked, and
in-order datagrams are placed into \textsf{\_appBuffers} for consumption by the
application thread.

Handling a smooth packet flow is critical for performance. RSP actively flow
control to advance the stream. Each incoming connection actively acknowledges
every $n$ (17 by default) packets fully received. The incoming connections shift
this acknowledgment by their connection identifier to avoid bursts of acks. Any
missed datagram is actively nack'ed as soon as it is detected. Write connections
continuously retransmit packets from nack datagrams, and advance their
\textsf{\_writeBuffers} upon reception of acks from all readers. Furthermore,
the writer will explicitly request an ack or nack when it runs out of empty
buffers or finishes its write queue.

Congestion control is necessary to optimize bandwidth usage. While TCP uses the
well-known AIMD (additive increase, multiplicative decrease) algorithm, we have
chosen a more aggressive congestion control algorithm of additive increase and
decrease. This has proven experimentally to be more optimal: UDP is often
rate-limited by switches, that is, packets are discarded regularly and not
exceptionally. Only slowly backing of the current send rate helps to stay close
to this limit. Furthermore, our RSP traffic is limited to the local subnet,
making cooparation between multiple data stream less of an issue. Send rate
limiting uses a bucket algorithm, where over time the bucket fills with send
credits, from which sends are substracted. If there are no available credits,
the sender sleeps until sufficient credits are available.

@@

Adapting an existing application for parallel rendering requires the
synchronization of application data across the processes in the parallel
rendering setup. Existing parallel rendering frameworks address this often
poorly, at best they rely on MPI to distribute data. Real-world, interactive
visualization applications are typically written in C++ and have complex data
models and class hierarchies to represent their application state. As outlined
in \cite{EMP:09}, the parallel rendering code in an Equalizer application only
needs access to the data needed for rendering, as all application logic is
centralized in the application main thread. We have encountered two main
approaches to address this distribution: Using a shared filesystem for static
data, or using data distribution for static and dynamic data.

Distributed objects are not required to build Equalizer applications. While most
developers choose to use this abstraction for convenience, we have seen
applications using other means for data distribution, e.g., MPI.

\subsubsection{Programming Interface}

Distributed objects in Collage provide powerful, object-oriented data
distribution for C++ objects. They facilitate the implementation of data
distribution in a cluster environment. Distributed objects are created by
subclassing from \textsf{co::Serializable} or \textsf{co::Object}. The
application programmer implements serialization and deserialization of the
distributed data. Distributed objects can be static (immutable) or
dynamic. Objects have a universally unique identifier (UUID) to address them
cluster-wide. A master-slave model is used to establish mapping and data
synchronization across processes. Typically, the application main loop registers
a master instance and communicates the UUID to the render clients, which map
their instance to the given identifier. The following object types are
available:

@@

Besides the instance data serialization methods used to map an object, delta and
unbuffered objects may implement \textsf{pack} and \textsf{unpack} to serialize
or deserialize the changes since the last version.

@@

\subsubsection{Optimizations}

The API presented in the previous section provides sufficient abstraction to
implement various optimizations for faster mapping and synchronization of data:
compression, chunking, caching, preloading and multicast. The results section
evaluates these optimizations.

The most obvious one is compression. Recently, many new compression algorithms
have been developed which exploit modern CPU architectures and deliver
compression rates well above one gigabyte per second. Collage uses the Pression
library~\cite{pression}, which provides an unified interface for a number of
compression libraries, such as FastLZ~\cite{fastlz}, Snappy~\cite{snappy} and
ZStandard~\cite{zstd}. It also contains a custom, virtually zero-cost RLE
compressor. Pression also parallelizes the compression and decompression using
data decomposition.

The data streaming interface also implements chunking, which aims to pipeline
the serialization code with the network transmission. After a configurable
number of bytes has been serialized to the internal buffer, it is transmitted
and serialization continues. This is used both for the initial mapping data and
for commit data.

Caching retains instance data of objects in a cache, and reuses this data to
accelarate mapping of objects. The instance cache is either filled by "snooping"
on multicast transmissions or by an explicit preloading when the master objects
are registered. The preloading ("send on register") sends instance data of
recently registered master objects while the correspinding node is idle to all
connected nodes. These nodes simply enter the received data to their
cache. Preloading uses multicast when available.

Due to the master-slave model of data distribution, multicast can be used to
optimize the transmission time of data. If the contributing nodes share a
multicast session, and more than one slave instance is mapped, Collage
automatically uses the multicast connection to send the new version
information.

@@

\subsection{Raster}

\subsection{Bino}

\subsection{Omegalib}

@@

Pression data compressor benchmark

Distribute large ply and raw model to 1..16 nodes using TCP, IB and RSP
